---
title: "reddisco-quarto"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
---

## RNA Editing Quarto Notebook (Downstream Analysis)

This notebook processes per-sample, per-replicate RNA editing output from RediTools v1
and replicates the manual filtering workflow in a reproducible, scriptable form.

The workflow for each cell line:

1. **Recovery step** — for each replicate, sites in the master annotation set that fell
   below the upstream 10% frequency filter are recovered from the pre-filter file and
   re-attached with their actual FREQ/COV values (or 0 if truly absent).
2. **List 1(a/b)** — intersect replicates within each condition (sites with FREQ > 0 in
   both replicates); compute per-site mean FREQ.
3. **List 2(a)** — intersect siNT and sip150 consistent sites; compute
   Δ = F(mean)siNT − F(mean)sip150.
4. **List 2(b)** — sites present in siNT (List 1a) but absent from sip150 (List 1b).
5. **List 3** — union of List 2(a) with positive Δ and List 2(b): final editing-reduction
   sites.

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(fs)
```

---

## Configuration

```{r config}
# Root directory containing all TSV files, organised by cell line.
# Expected structure:
#   data/
#     ORL-207/
#       ORL-207-siNT-R1.known_sites_freq10pct_genic.tsv     <- annotated (SnpEff)
#       ORL-207-siNT-R1-knownEditing-labeled.tsv             <- pre-filter (no SnpEff)
#       ... (same pattern for R2, sip150-R1, sip150-R2)
#     SCC9/  ...
#     BICR10/  ...  (ensure R1 is the corrected, relabelled file)
DATA_DIR <- "data"

# Cell lines to process
CELL_LINES <- c("ORL-207", "SCC9", "BICR10")

# File suffix for the SnpEff-annotated, frequency-filtered files
ANNOTATED_SUFFIX <- "known_sites_freq10pct_genic.tsv"

# File suffix for the pre-filter files (AG-only, no frequency cutoff, NO header)
# Adjust this to match your actual pre-filter filename suffix
PREFILTER_SUFFIX <- "known_sites_AG_genic.tsv"

# Columns used to uniquely identify an editing site across files
SITE_KEYS <- c("CHROM", "POS")

# Column names for the headerless pre-filter file (RediTools AG-filtered output)
PREFILTER_COLS <- c(
  "CHROM", "POS", "REF", "STRAND", "COV", "MeanQ",
  "BaseCount", "AllSubs", "FREQ",
  "gCOV", "gMeanQ", "gBaseCount", "gAllSubs", "gFREQ",
  "REPTYPE", "REPNAME", "SNPFlag", "dbSNP_ID",
  "REDIPortal", "EDITSTATUS"
)
```

---

## Helper functions

### File readers

```{r fn-readers}
# Read a SnpEff-annotated, frequency-filtered TSV and rename awkward ANN columns
read_annotated_tsv <- function(path) {
  read_tsv(path, show_col_types = FALSE) |>
    rename(
      gene    = `ANN[0].GENE`,
      feature = `ANN[0].FEATUREID`,
      effect  = `ANN[0].EFFECT`,
      impact  = `ANN[0].IMPACT`,
      biotype = `ANN[0].BIOTYPE`,
      hgvs_c  = `ANN[0].HGVS_C`
    )
}

# Read a headerless pre-filter TSV (RediTools AG-only output, no frequency cutoff)
read_prefilter_tsv <- function(path) {
  read_tsv(path, col_names = PREFILTER_COLS, show_col_types = FALSE)
}

# Build a file path given cell line, condition, replicate, and suffix
build_path <- function(cell_line, condition, replicate, suffix) {
  path(DATA_DIR, cell_line,
       paste0(cell_line, "-", condition, "-", replicate, ".", suffix))
}
```

### Recovery step

```{r fn-recovery}
# Build a master annotation table from all four annotated files for a cell line.
# Keeps the first-seen annotation for any site that appears in multiple files.
build_master_annotation <- function(cell_line) {
  combos <- expand_grid(
    condition = c("siNT", "sip150"),
    replicate = c("R1", "R2")
  )

  map2(combos$condition, combos$replicate, function(cond, rep) {
    p <- build_path(cell_line, cond, rep, ANNOTATED_SUFFIX)
    if (file_exists(p)) read_annotated_tsv(p) else NULL
  }) |>
    compact() |>
    bind_rows() |>
    distinct(CHROM, POS, .keep_all = TRUE)
}

# For one replicate: recover sites in the master annotation set that were dropped
# by the upstream frequency filter.  Returns the annotated dataframe with dropped
# sites re-attached (FREQ/COV from the pre-filter file, or 0 if truly absent).
# A `recovered` column flags which rows were added this way.
recover_replicate <- function(annotated_df, prefilter_path, master_annot) {

  # Sites in the master that are absent from this replicate's annotated file
  missing <- anti_join(master_annot, annotated_df, by = SITE_KEYS)

  if (nrow(missing) == 0) {
    return(mutate(annotated_df, recovered = FALSE))
  }

  # Pull FREQ and COV for missing sites from the pre-filter file
  prefilter <- read_prefilter_tsv(prefilter_path) |>
    select(CHROM, POS, FREQ, COV)

  recovered_rows <- missing |>
    # Keep annotation columns only (FREQ/COV will come from pre-filter)
    select(CHROM, POS, REF, ALT, STRAND,
           gene, feature, effect, impact, biotype, hgvs_c,
           REPTYPE, REPNAME, EDITSTATUS) |>
    left_join(prefilter, by = SITE_KEYS) |>
    mutate(
      FREQ      = replace_na(FREQ, 0),   # 0 = truly absent in this replicate
      COV       = replace_na(COV,  0),
      recovered = TRUE
    )

  bind_rows(
    mutate(annotated_df, recovered = FALSE),
    recovered_rows
  )
}

# Load and recover one replicate: wrapper that calls both readers and recover_replicate
load_replicate <- function(cell_line, condition, replicate, master_annot) {
  ann_path <- build_path(cell_line, condition, replicate, ANNOTATED_SUFFIX)
  pre_path <- build_path(cell_line, condition, replicate, PREFILTER_SUFFIX)

  annotated <- read_annotated_tsv(ann_path)
  recover_replicate(annotated, pre_path, master_annot)
}
```

### List-building functions

```{r fn-lists}
# List 1: intersect two replicates for one condition.
# "Consistent" = site has FREQ > 0 in BOTH replicates after recovery.
# Returns per-site mean FREQ with individual replicate FREQs retained.
make_list1 <- function(r1_df, r2_df, condition_label) {
  inner_join(
    r1_df |> select(all_of(SITE_KEYS), gene, effect, impact, biotype,
                    REPTYPE, REPNAME, EDITSTATUS, STRAND,
                    freq_R1 = FREQ, cov_R1 = COV, recovered_R1 = recovered),
    r2_df |> select(all_of(SITE_KEYS),
                    freq_R2 = FREQ, cov_R2 = COV, recovered_R2 = recovered),
    by = SITE_KEYS
  ) |>
    filter(freq_R1 > 0, freq_R2 > 0) |>   # both replicates must show evidence
    mutate(
      condition    = condition_label,
      mean_freq    = (freq_R1 + freq_R2) / 2,
      mean_cov     = (cov_R1  + cov_R2)  / 2,
      any_recovered = recovered_R1 | recovered_R2
    )
}

# List 2(a): sites consistent in BOTH conditions.
# Computes delta = F(mean)siNT - F(mean)sip150.
make_list2a <- function(list1a, list1b) {
  inner_join(
    list1a |> select(all_of(SITE_KEYS), gene, effect, impact, biotype,
                     REPTYPE, REPNAME, EDITSTATUS, STRAND,
                     mean_freq_siNT   = mean_freq,
                     mean_cov_siNT    = mean_cov,
                     freq_R1_siNT     = freq_R1,
                     freq_R2_siNT     = freq_R2,
                     recovered_siNT   = any_recovered),
    list1b |> select(all_of(SITE_KEYS),
                     mean_freq_sip150 = mean_freq,
                     mean_cov_sip150  = mean_cov,
                     freq_R1_sip150   = freq_R1,
                     freq_R2_sip150   = freq_R2,
                     recovered_sip150 = any_recovered),
    by = SITE_KEYS
  ) |>
    mutate(
      delta_freq  = mean_freq_siNT - mean_freq_sip150,
      list_origin = "2a_shared"
    )
}

# List 2(b): sites in siNT only (absent from sip150 consistent list).
make_list2b <- function(list1a, list1b) {
  anti_join(list1a, list1b, by = SITE_KEYS) |>
    mutate(
      delta_freq  = mean_freq,   # full siNT frequency; sip150 is absent/0
      list_origin = "2b_siNT_only"
    )
}

# List 3: editing-reduction sites.
# List 2(a) with positive delta  +  all List 2(b).
make_list3 <- function(list2a, list2b) {
  bind_rows(
    filter(list2a, delta_freq > 0),
    list2b
  ) |>
    arrange(desc(delta_freq))
}

# Master function: run the full workflow for one cell line
run_workflow <- function(cell_line) {
  message("\n── Processing: ", cell_line, " ──")

  # Build master annotation from all four annotated files
  master <- build_master_annotation(cell_line)
  message("  Master annotation set: ", nrow(master), " sites")

  # Load and recover each replicate
  siNT_R1   <- load_replicate(cell_line, "siNT",   "R1", master)
  siNT_R2   <- load_replicate(cell_line, "siNT",   "R2", master)
  sip150_R1 <- load_replicate(cell_line, "sip150", "R1", master)
  sip150_R2 <- load_replicate(cell_line, "sip150", "R2", master)

  message("  Recovered sites — siNT-R1: ",   sum(siNT_R1$recovered),
          "  siNT-R2: ",                      sum(siNT_R2$recovered),
          "  sip150-R1: ",                    sum(sip150_R1$recovered),
          "  sip150-R2: ",                    sum(sip150_R2$recovered))

  list1a <- make_list1(siNT_R1,   siNT_R2,   "siNT")
  list1b <- make_list1(sip150_R1, sip150_R2, "sip150")
  list2a <- make_list2a(list1a, list1b)
  list2b <- make_list2b(list1a, list1b)
  list3  <- make_list3(list2a, list2b)

  list(
    cell_line = cell_line,
    list1a    = list1a,
    list1b    = list1b,
    list2a    = list2a,
    list2b    = list2b,
    list3     = list3
  )
}
```

---

## Run workflow for all cell lines

```{r run-all, message=TRUE}
results <- set_names(
  map(CELL_LINES, run_workflow),
  CELL_LINES
)
```

---

## Summary table

```{r summary-table}
map_dfr(results, function(r) {
  tibble(
    cell_line               = r$cell_line,
    n_siNT_consistent       = nrow(r$list1a),
    n_sip150_consistent     = nrow(r$list1b),
    n_list2a_shared         = nrow(r$list2a),
    n_list2b_siNT_only      = nrow(r$list2b),
    n_list3_final           = nrow(r$list3),
    # How many final sites contain at least one recovered replicate value
    n_list3_any_recovered   = sum(r$list3$any_recovered, na.rm = TRUE)
  )
})
```

---

## Visualisation

### Delta frequency distribution (List 2a)

```{r plot-delta, fig.width=9, fig.height=4, warning=FALSE}
map_dfr(results, function(r) {
  r$list2a |> mutate(cell_line = r$cell_line)
}) |>
  ggplot(aes(x = delta_freq, fill = cell_line)) +
  geom_histogram(binwidth = 0.02, colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~ cell_line, scales = "free_y") +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Distribution of editing frequency change (siNT − sip150)",
    subtitle = "Positive = editing reduced in sip150; negative = editing increased",
    x = "Δ editing frequency", y = "Number of sites"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

### Mean editing frequency: siNT vs sip150 (List 2a)

Recovered sites (where at least one replicate value came from the pre-filter file)
are shown with a different shape so they can be assessed separately.

```{r plot-scatter, fig.width=10, fig.height=4, warning=FALSE}
map_dfr(results, function(r) {
  r$list2a |> mutate(cell_line = r$cell_line)
}) |>
  ggplot(aes(x = mean_freq_siNT, y = mean_freq_sip150,
             colour = delta_freq > 0,
             shape  = recovered_siNT | recovered_sip150)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey40") +
  facet_wrap(~ cell_line) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_manual(
    values = c("TRUE" = "#e05c5c", "FALSE" = "#5c8de0"),
    labels = c("TRUE" = "Reduced", "FALSE" = "Not reduced")
  ) +
  scale_shape_manual(
    values = c("TRUE" = 2, "FALSE" = 16),
    labels = c("TRUE" = "Recovered", "FALSE" = "Primary call")
  ) +
  labs(
    title  = "Mean editing frequency: siNT vs sip150",
    x      = "F(mean) siNT",
    y      = "F(mean) sip150",
    colour = "Editing reduced",
    shape  = "Data origin"
  ) +
  theme_bw()
```

### List 3 composition: shared vs siNT-only

```{r plot-list3-composition, fig.width=6, fig.height=4}
map_dfr(results, function(r) {
  r$list3 |> mutate(cell_line = r$cell_line)
}) |>
  count(cell_line, list_origin) |>
  ggplot(aes(x = cell_line, y = n, fill = list_origin)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    values = c("2a_shared" = "#e05c5c", "2b_siNT_only" = "#e09a5c"),
    labels = c("2a_shared" = "Shared – reduced", "2b_siNT_only" = "siNT only – abolished")
  ) +
  labs(
    title = "List 3: final editing-reduction sites by origin",
    x = NULL, y = "Number of sites", fill = NULL
  ) +
  theme_bw()
```

### Repeat type breakdown (List 3)

```{r plot-reptype, fig.width=8, fig.height=4}
map_dfr(results, function(r) {
  r$list3 |> mutate(cell_line = r$cell_line)
}) |>
  mutate(REPTYPE = replace_na(REPTYPE, "Non-repetitive")) |>
  count(cell_line, REPTYPE) |>
  ggplot(aes(x = cell_line, y = n, fill = REPTYPE)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Repeat type composition of final editing-reduction sites",
    x = NULL, y = "Proportion", fill = "Repeat type"
  ) +
  theme_bw()
```

---

## Export

```{r export, eval=FALSE}
out_dir <- "output/reddisco"
dir_create(out_dir)

walk(results, function(r) {
  cl <- r$cell_line
  write_tsv(r$list1a, path(out_dir, paste0(cl, "_list1a_siNT_consistent.tsv")))
  write_tsv(r$list1b, path(out_dir, paste0(cl, "_list1b_sip150_consistent.tsv")))
  write_tsv(r$list2a, path(out_dir, paste0(cl, "_list2a_shared.tsv")))
  write_tsv(r$list2b, path(out_dir, paste0(cl, "_list2b_siNT_only.tsv")))
  write_tsv(r$list3,  path(out_dir, paste0(cl, "_list3_final_reduction.tsv")))
  message("Exported: ", cl)
})
```

